How to select predictive models for decision-making or causal inference?

### 2024-08-25 how_to_select_latex.tex
WIP: Reviewer1

Should we know the two expected outcomes for a given $X$,
we could compute the
difference between them, which gives the causal effect of the treatment.
%
These two expected outcomes can be estimated from observed data:
the consistency \ref{assumption:consistency} and unconfoundedness
\ref{assumption:ignorability} assumptions imply the following equality:
\begin{equation}\label{eq:mu_identification}
    \mathbb E_{Y(a) \sim \mathcal{D^{\star}}} [Y(a)|X=x] = \mathbb E_{Y \sim \mathcal{D}} [Y|X=x, A=a]
\end{equation}
On the left, the expectation is taken on the counterfactual unobserved
distribution. On the right, the expectation is taken on the factual observed
distribution conditionally on the treatment. For the rest of the
paper, the expectations will always be taken on the factual observed
distribution $\mathcal{D}$. This identification leads to outcome based estimators (ie.
g-computation estimators \cite{snowden_implementation_2011}):
\begin{eqnarray}
    \tau =& \mathbb E_{Y \sim \mathcal{D^{\star}}}[Y(1) - Y(0)|X=x]
    \notag
    \\
    = &\mathbb E_{Y \sim \mathcal{D}}[Y|A=1] - \mathbb E_{Y \sim \mathcal{D}}[Y| A=0]
    \label{eq:tau_population}
\end{eqnarray}
This equation builds on two quantities: the conditional expectancy
of the outcome given the covariates and either
treatment or no no treatment, called \emph{response function}:
\begin{flalign*}
    \text{Response function}
     &  &
    \mu_{a}(x) \myeq \; \mathbb E_{Y \sim \mathcal{D}} [Y|X=x, A=a]
     &  &
\end{flalign*}

Given a sample of data and the oracle response functions $\mu_0, \mu_1$, the
finite sum version of \autoref{eq:tau_population} leads to an
estimator of the ATE written:
\begin{equation}
    \hat \tau = \frac{1}{n} \biggl(\sum_{i=1}^n \mu_{1}(x_i) - \mu_{0}(x_i) \biggr)
    \label{eq:ate_estimate}
\end{equation}
This estimator is an oracle \emph{finite sum} estimator by opposition to the
population expression of $\tau$, $\mathbb{E}[\mu_{1}(x_i) - \mu_{0}(x_i)]
$,
which involves an expectation taken on the full
distribution $\mathcal D$, which is observable but requires infinite data. For
each estimator $\ell$ taking an expectation over $\mathcal D$, we use the symbol
$\hat \ell$ to note its finite sum version.

@@

\paragraph{Robinson decomposition}
The \emph{R-decomposition}
of the outcome model\cite{robinson_rootnconsistent_1988} plays an important role:
introducing two quantities, the conditional mean outcome
and the probability to be treated (known as propensity score \cite{rosenbaum_central_1983}):
\begin{flalign}
    \text{Conditional mean outcome} &                    & m(x) \myeq \; & \mathbb E_{Y \sim
    \mathcal{D}} [Y|X=x]            &                    &
    \label{def:m}
    \\
    \text{Propensity score}         &                    &
    e(x) \myeq \;                   & \mathbb P[A=1|X=x]
    \label{def:propensity_score}
\end{flalign}
the outcome can be written
\begin{flalign}\label{eq:r_decomposition}
    \text{R-decomposition}            &  & y(a) = m(x) + \big( a - e(x) \big)
    \tau(x) + \varepsilon(x; a)\notag &  &
    \\\ &&\text{with}\quad \mathbb E[\varepsilon(X; A)|X, A] = 0&&
\end{flalign}
$m$ and $e$ are often called
\emph{nuisances} \cite{chernozhukov_double_2018}; they are unknown.

>>>>

Should we know the two expected outcomes for a given $X$,
we could compute the
difference between them, which gives the causal effect of the treatment.
%
These two expected outcomes can be estimated from observed data:
the consistency \ref{assumption:consistency} and unconfoundedness
\ref{assumption:ignorability} assumptions imply the following equality:
\begin{equation}\label{eq:mu_identification}
    \mathbb E_{Y(a) \sim \mathcal{D^{\star}}} [Y(a)|X=x] = \mathbb E_{Y \sim \mathcal{D}} [Y|X=x, A=a]
\end{equation}
On the left, the expectation is taken on the counterfactual unobserved
distribution. On the right, the expectation is taken on the factual observed
distribution conditionally on the treatment. For the rest of the
paper, the expectations will always be taken on the factual observed
distribution $\mathcal{D}$. This identification leads to outcome based estimators (ie.
g-computation estimators \cite{snowden_implementation_2011}):
\begin{eqnarray}
    \tau =& \mathbb E_{Y \sim \mathcal{D^{\star}}}[Y(1) - Y(0)]
    \notag
    \\
    = &\mathbb E_{Y \sim \mathcal{D}}[Y|A=1] - \mathbb E_{Y \sim \mathcal{D}}[Y| A=0]
    \label{eq:tau_population}
\end{eqnarray}
This equation builds on the conditional expectation of the outcome given the treatment $\mathbb E_{\sim \mathcal{D}}[Y|A]$. Outcome based methods target this quantity conditionally on the covariates, called \emph{response function}:
\begin{flalign*}
    \text{Response function}
     &  &
    \mu_{a}(x) \myeq \; \mathbb E_{Y \sim \mathcal{D}} [Y|X=x, A=a]
     &  &
\end{flalign*}

Given a sample of data and the oracle response functions $\mu_0, \mu_1$, the
finite sum version of \autoref{eq:tau_population} leads to an unbiased
estimator of the ATE written:
\begin{equation}
    \hat \tau = \frac{1}{n} \biggl(\sum_{i=1}^n \mu_{1}(x_i) - \mu_{0}(x_i) \biggr)
    \label{eq:ate_estimate}
\end{equation}
This estimator is an oracle \emph{finite sum} estimator by opposition to the
population expression of $\tau$, $\mathbb{E}[\mu_{1}(x_i) - \mu_{0}(x_i)]
$,
which involves an expectation taken on the full
distribution $\mathcal D$, which is observable but requires infinite data. For
each estimator $\ell$ taking an expectation over $\mathcal D$, we use the symbol
$\hat \ell$ to note its finite sum version. In practice the response function can be estimated with one statistical model taking the treatment as a supplementary covariate $\hat \mu(x, a)$ (called S-learner) or by two separate model fitted on each population, treated or not $(\hat \mu_0(x), \hat \mu_1(x))$ (T-learner) \cite{kunzel_metalearners_2019}.

@@

%TODO: not well located here. The reader does not know why we introduce this decomoositon her.
\paragraph{Robinson decomposition}
The \emph{R-decomposition}
of the outcome model\cite{robinson_rootnconsistent_1988} links the counterfactual outcomes $y(a)$ to the CATE by introducing two quantities, the conditional mean outcome
and the probability to be treated (known as propensity score \cite{rosenbaum_central_1983}).:
\begin{flalign}
    \text{Conditional mean outcome} &                    & m(x) \myeq \; & \mathbb E_{Y \sim
    \mathcal{D}} [Y|X=x]            &                    &
    \label{def:m}
    \\
    \text{Propensity score}         &                    &
    e(x) \myeq \;                   & \mathbb P[A=1|X=x]
    \label{def:propensity_score}
\end{flalign}
the outcome can be written
\begin{flalign}\label{eq:r_decomposition}
    \text{R-decomposition}            &  & y(a) = m(x) + \big( a - e(x) \big)
    \tau(x) + \varepsilon(x; a)\notag &  &
    \\\ &&\text{with}\quad \mathbb E[\varepsilon(X; A)|X, A] = 0&&
\end{flalign}
$m$ and $e$ are often called
\emph{nuisances} \cite{chernozhukov_double_2018}; they are unknown and are estimated in practice thanks to statistical models.
